---
title: "Getting started with flowr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with flowr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`flowr` brings [flowR](https://github.com/flowr-analysis/flowr) --- a static
dataflow analyzer and program slicer for R --- to your R session. It lets you
*slice* a script down to the statements that affect a given variable, extract a
script's *dependencies*, run flowR *queries*, and drive the flowR *REPL*, all
from R.

## Installation

```{r}
# install.packages("remotes")
remotes::install_github("flowr-analysis/flowr-r-adapter")
```

## The one-liner

Install a flowR engine once, then the high-level helpers open a shared session
the first time they are called, so the rest needs no setup:

```{r}
library(flowr)
flowr_install()   # once: downloads a self-contained flowR binary

code <- "x <- 1\ny <- 2\nz <- x + 5\nprint(z)"

# Which statements does `z` on line 3 depend on?
cat(slice(code, "3@z")$code)
```

A slicing *criterion* is written as `line@variable` (`"3@z"`), `line:column`
(`"3:1"`) or `$id` for a flowR node id. Pass several at once with a vector.

`slice()` returns a `flowr_slice`. Its **`$code`** element is the reconstructed,
sliced R source as a single string - this is what you pass on, `cat()`, or write
to a file (`writeLines(slice(code, "3@z")$code, "sliced.R")`). `$lines` are the
kept original line numbers. **Printing** the slice instead shows a colored view:
by default a git-style diff (removed lines red with `-`, kept lines as context),
`print(s, style = "gray")` dims the unused lines, and `print(s, style = "code")`
prints just `$code`. Slice a whole project with
`slice(folder = "path/to/project", criterion = ...)`.

## Diagnostics

```{r}
flowr_status()               # engine, configuration and active session
options(flowr.debug = TRUE)  # log engine start-up and protocol messages (R side)
flowr_set_config(verbose = TRUE)  # run the flowR server with --verbose
flowr_log()                  # read the flowR server's own log output
flowr_bug_report()           # open a prefilled GitHub issue (uses your browser)
flowr_feedback()             # share feedback
```

## Overview: what a script does

`flowr_overview()` is the convenience summary: the libraries loaded, files
sourced, data read/written and plots produced, printed nicely. Every item carries
a slicing `criterion`, so you can slice straight onto it - for example, see
exactly the code behind the first plot:

```{r}
code <- "library(dplyr)\nd <- read.csv('in.csv')\nplot(d)"
ov <- flowr_overview(code)
ov                                              # a formatted overview
cat(slice(code, ov$visualize[[1]]$criterion)$code)
```

The full query API is `query()` (`dataflow`, `dependencies`, `linter`, and more -
see `vignette("flowr-engines")`).

## Inspect a whole project

Inside an R package or project you usually do not even pass code: the
project-aware commands default to the current project. `flowr` walks up to the
outermost project root (a `DESCRIPTION` or `.Rproj`, else a git root, else a
folder with `.R` files) and reports what it picked.

```{r}
setwd("path/to/a/package")

inspect_project()   # files and their roles (source, test, vignette, data, ...)
flowr_overview()    # the package's libraries, plots, reads/writes - ready to slice
```

`inspect_project()`, `flowr_overview()`, `slice()`, `query()` and friends all
accept `code`, `file` or `folder`; with none, they analyse the enclosing project.

## Engines, in one paragraph

`flowr` obtains flowR in several interchangeable ways --- the flowR JS+wasm
**bundle** shipped in the package (needs Node.js, no download), a self-contained
downloaded **binary** (no Node or Docker required), a system/private **Node.js**,
or **Docker**. flowR itself is always the analysis backend; there is no
approximate fallback. `auto` (the default) prefers a cached binary, then the
bundle when Node is present, then downloads the binary. To pre-install an engine:

```{r}
flowr_install()                  # a self-contained binary into the cache
flowr_install(engine = "docker") # or pull the flowR docker image
flowr_installed("all")           # which engines are ready right now?
```

See `vignette("flowr-engines")` for choosing and configuring engines, the query
API, the REPL, secure mode and performance.

## A native flowR shell

`flowr_repl()` relays single commands, but for an interactive session with
flowR's own tab completion and history, hand over the terminal with
`flowr_console()` (interactive `flowr_repl()` uses it automatically):

```{r}
flowr_repl(":dataflow x <- 1")   # one command, result back in R
flowr_console()                   # full flowR shell; :quit to return
flowr_console(r_access = TRUE)    # ... that can also execute R (uses r-shell)
```

In the console you can pass `watch://<path>` instead of `file://<path>` to a
`:`-command to re-run it whenever the file changes. From R, `flowr_watch()` does
the same for any command:

```{r}
flowr_watch("script.R")          # re-print the overview on every save
flowr_watch("script.R", function(f) print(slice(file = f, criterion = "8@out")))
```

## Configuration and updates

Read the effective settings with `flowr_config()` (grouped, with the server-only
knobs marked), change them with `flowr_set_config()`, and check for a newer flowR
with `flowr_update()`:

```{r}
flowr_config()                                  # the effective configuration
flowr_set_config(engine = "binary", request_timeout = 300)
flowr_set_config(flowr_version = "2.12.0", persist = TRUE)  # also across sessions
flowr_update()                                  # is a newer flowR available?
```

Every knob is also a plain R option, so `options(flowr.engine = "binary")` and
`cfg <- flowr_config(); cfg$engine <- "binary"` work too. Turn on
`options(flowr.timing = TRUE)` to print each command's wall-clock time, and
`options(flowr.debug = TRUE)` to log engine start-up and protocol messages.

## Managing sessions

You usually do not need to think about sessions --- a default one is created and
reused. When you want an explicit one (e.g. a specific engine), use
`flowr_connect()` and release it with `flowr_disconnect()`:

```{r}
s <- flowr_connect(engine = "binary")
slice(code, "4@z")
flowr_disconnect(s)
```
