---
title: "Binary integrity and signature verification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary integrity and signature verification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The `binary` engine is a self-contained flowR executable that `flowr` downloads
from GitHub on demand. Because it is code that will run on your machine, `flowr`
verifies it before it is ever executed. This vignette explains the two layers of
verification, what [flowr_status()] reports, and --- if you see *"checksum only
(no signature)"* --- exactly what to do to turn on signature validation.

## Two layers of protection

1. **SHA-256 checksum (always).** Every download is checked against a published
   `.sha256` sidecar. This detects corruption and truncation. It uses the
   `digest` package, which `flowr` depends on, so it is always available.
2. **Cryptographic signature (optional but recommended).** Release binaries are
   signed with the project's private key. `flowr` ships the matching **public
   key** (`inst/flowr-pubkey.pem`) and, when the `openssl` package is installed,
   verifies the downloaded archive's `.sig` against it. This proves the binary
   was produced by the flowR release pipeline, not just that it arrived intact.

`flowr` records how strongly each installed binary was verified, and
[flowr_status()] shows it on the `binary` line:

- `signature-verified` --- checksum **and** signature matched the pinned key.
- `checksum only (no signature)` --- the checksum matched but the signature step
  was skipped (see below).
- `unverified` --- neither could be recorded (an old or hand-placed binary).

## Turning on signature validation

If `flowr_status()` shows *checksum only*, do this:

```{r}
# 1. install the verifier (openssl is a suggested, not required, dependency)
install.packages("openssl")

# 2. keep verification on (both are on by default)
options(flowr.secure = TRUE)          # or flowr_set_config(verify_signature = TRUE)

# 3. re-download the binary so verification runs on a fresh copy
flowr_install(engine = "binary", force = TRUE)

flowr_status()                        # should now read: signature-verified
```

The verification level is recorded **at install time**, so simply installing
`openssl` afterwards does not re-verify a binary that is already on disk --- the
`force = TRUE` reinstall is what makes `flowr` fetch the `.sig` and check it.

## Why you might see "checksum only"

- **`openssl` is not installed.** It is the verifier; without it the signature
  step is skipped (in the default *secure* mode a fresh install would instead
  stop and ask you to install it).
- **No signature is published** for that flowR version/platform yet, so there is
  nothing to verify against.
- **Verification was turned off** with `options(flowr.secure = FALSE)` or
  `flowr_set_config(verify_signature = FALSE)`.

## Secure mode

`secure` mode is on by default (`options(flowr.secure = TRUE)`). In secure mode a
binary that cannot be verified is **refused**, not silently downgraded: if a
public key is pinned but `openssl` is missing, or no signature is available, the
install stops with an explanation rather than running an unverified executable.
Turning secure mode off relaxes this to a best-effort checksum, which is why an
unsigned binary can then install as *checksum only*.

## Not downloading a binary at all

If you would rather not download and verify an executable, use the JS + WebAssembly
**bundle** that ships inside the package (it needs Node.js but no download):

```{r}
flowr_connect(engine = "bundled")
```

See `vignette("flowr-engines")` for the full engine picture, and [flowr_status()]
/ [flowr_config()] for what is currently in effect.
