---
title: "Engines, queries, the REPL and secure mode"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engines, queries, the REPL and secure mode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(flowr)
```

This vignette covers how `flowr` obtains flowR, how to pick and configure an
engine, and how to use the query API and the REPL.

## Engines

An *engine* is how a running flowR server is obtained. All engines are
interchangeable; the client speaks the same protocol to each. flowR is always
the analysis backend; there is no approximate fallback.

| Engine    | Needs                       | Notes |
|-----------|-----------------------------|-------|
| `bundled` | Node.js                     | flowR JS+wasm shipped in the package; no download. |
| `binary`  | nothing (downloaded once)   | Self-contained executable; no Node needed. |
| `node`    | Node.js (system or private) | npm-installs the flowR package. |
| `docker`  | Docker                      | Runs `eagleoutice/flowr`. |
| `auto`    | --                          | Cached binary, else the shipped bundle (Node), else the binary. |

Select an engine per call, or globally:

```{r}
library(flowr)

flowr_connect(engine = "binary")            # per session
options(flowr.engine = "node")              # globally
Sys.setenv(FLOWR_ENGINE = "docker")         # via environment
```

`auto` (the default) uses an already-installed engine, otherwise it offers to
install the self-contained binary. Install an engine explicitly:

```{r}
flowr_install()                    # binary engine, flowR 2.11.1
flowr_install(engine = "node")     # Node engine
flowr_is_installed()               # TRUE / FALSE
flowr_uninstall()                  # clear the cache
```

Downloads are cached under `tools::R_user_dir("flowr", "cache")` and, for the
binary engine, verified against a SHA-256 manifest before ever being executed.

### The flowR parser engine

Independently of the above, flowR itself can parse R with **tree-sitter** (the
default; needs no R) or the **r-shell** engine, which reuses the R you are
already running:

```{r}
flowr_connect(flowr_engine = "tree-sitter")   # default
flowr_set_config(secure = FALSE)               # r-shell needs secure mode off
flowr_connect(flowr_engine = "r-shell")        # flowR parses via your R
```

Because you are inside R, enabling R *execution* in flowR's REPL is a one-liner:
`flowr_console(r_access = TRUE)` starts the flowR shell with the r-shell engine
and `--r-session-access`, so `:`-commands and plain R both work.

## Secure mode (on by default)

`flowr` runs in a hardened mode unless you opt out. Secure mode:

* forces the tree-sitter parser, so untrusted code is never handed to an R
  interpreter;
* always checksum-verifies binary downloads;
* never invokes a shell (processes are started from an argument vector, so there
  is no shell-escape surface).

```{r, eval = TRUE}
flowr_config()$secure              # TRUE by default
```

```{r}
flowr_set_config(secure = FALSE)   # opt out (e.g. to use the r-shell engine)
```

## The query API

`query()` exposes flowR's full query API. Pass a query type name, a query object
(a list with a `type`), or a list of them:

```{r}
code <- "x <- 1\ny <- x + 1\nprint(y)"

query(code, "dependencies")
query(code, "dataflow")
query(code, c("normalized-ast", "id-map"))

# object form for queries that take arguments
query(code, list(type = "static-slice", criteria = "2@y", direction = "backward"))
```

Supported query types in flowR 2.11.1 include `dependencies`, `dataflow`,
`call-context`, `df-shape`, `static-slice`, `id-map`, `normalized-ast`,
`linter`, `location-map`, `call-graph` and more. Because `query()` forwards
whatever you pass, new query types work without updating this package.

## Linting

`flowr_lint()` wraps the `linter` query (see `vignette("flowr")` for the tidy,
lintr-compatible output and the SARIF/GitHub/jarl formats). By default it runs
flowR's full rule set --- the same rules the flowR VS Code extension enables ---
and you can narrow that set, either per call or as a persistent default.

```{r}
# per call
flowr_lint("setwd('/x')\nx = 1", rules = c("absolute-file-paths", "naming-convention"))

# as a default for the session (or persist = TRUE for future sessions)
flowr_set_config(lint_rules = c("absolute-file-paths", "unused-definitions"))
flowr_lint("setwd('/x')")          # now runs only those two rules
flowr_set_config(lint_rules = character(0))   # back to flowR's full set
```

The rules available in flowR 2.11.1 are: `absolute-file-paths`,
`file-path-validity`, `seeded-randomness`, `deprecated-functions`,
`network-functions`, `dataframe-access-validation`, `dead-code`, `useless-loop`,
`problematic-inputs`, `stop-call`, `unused-definitions`, `naming-convention`,
`roxygen-arguments`, `no-leaked-credentials`, `undefined-symbol`,
`software-has-license` and `software-has-tests`. Each is documented on the flowR
wiki (e.g. <https://github.com/flowr-analysis/flowr/wiki/%5BLinting-Rule%5D-Absolute-File-Paths>);
`flowr_lint()`'s printed report links every rule to its page. Findings marked
`[fix]` can be applied with `flowr_lint_fix()`.

## The REPL

```{r}
flowr_repl(":help")                # a REPL command; returns the streamed output
flowr_repl("1 + 1")                # or an R expression
```

## Status and debugging

`flowr_status()` shows which engine you are using, the effective configuration,
the config file in effect and the active session. Set `options(flowr.debug =
TRUE)` to log engine start-up and every protocol message.

```{r, eval = TRUE}
flowr_status()
```

```{r}
options(flowr.debug = TRUE)        # log engine + protocol activity
slice("x <- 1\ny <- x", "2@y")
options(flowr.debug = FALSE)
```

Settings can also come from a `flowr.json` file (keys are option names, e.g.
`{"engine": "binary", "secure": true}`), resolved from
`options(flowr.config_file=)`, then `$FLOWR_CONFIG`, then `flowr.json` in the
working directory, then a per-user config file. `flowr_status()` shows which
file is active.

## Performance

A session keeps one warm flowR server and reuses it across calls, and analyses
are cached per content (bounded by `options(flowr.cache_size = )`), so repeated
`slice()`/`query()` on the same code avoid both re-spawning and re-analysis:

```{r}
s <- flowr_connect()
slice(code, "2@y")    # analyses once
query(code, "dataflow")  # reuses the cached analysis
flowr_clear_cache()               # if you need to drop it
flowr_disconnect(s)
```
